Characterizing System Dynamics using Diffusion Maps: A Chemotaxis Case Study

### 2014-06-06 chemotaxis/prl_paper.tex
started workong on prl edits
edited prl paper some more

TODO: rewrite according to the first 3 paragraphs of the introduction
%
Using a stochastic model of particle dynamics which arises in cellular chemotaxis, we illustrate how diffusion maps, a manifold learning technique, can be used to characterize the behavior of a dynamical system.
%
In our model, depending on the value of a single parameter, the macroscopic system dynamics range from heat equation dynamics, where particles diffuse uniformly outwards, to wave equation dynamics, where particle propagate with a fixed speed. 
%
We demonstrate that diffusion maps with the appropriate observers and metric can recover the two governing macroscopic system variables (initial condition and time) from data collected at the microscale. % in a variable- and equation-free manner. 
%
Furthermore, the order in which these variables are recovered is indicative of the system dynamics over the timescales of interest, 
and the eigenvalue spectrum indicates where this switch in dynamical behavior occurs.
%
%We demonstrate that an essential component of our analysis is working with the right observers (histograms/ensembles, rather than the single position of a particle) and with the right metric between those observers (earth mover's distance, rather than Euclidean distance).
%
%In heat equation mode, we find that time is more important than the initial conditions, as the particles quickly equilibrate.
%
%In wave equation mode, we find that the initial conditions are more important, as the initial distribution simply propagates in time.
%	
%Furthermore, these two components have a completely different ``nature". 
%
%One parameter (the initial condition) is governing the entire trajectory and the other parameter (time) is governing the ``step­--by­--step" evolution within each trajectory.

>>>>

Many dynamical systems are governed by stochastic, noisy motions at the microscale which give rise to smooth, coherent dynamics at the macroscale.
%
The relationship between the microscopic and macroscopic dynamics is often not immediately obvious and either requires some modeling assumptions or extensive observation of the system.
%
We demonstrate how extracting macroscopic dynamics from microscale data can be accomplished automatically using data-driven algorithms. 
%
These algorithms need no {\em a priori} knowledge about the appropriate macroscopic model, but instead, only require the definition of the appropriate observers and metric in order to recover the governing variables.
%
We illustrate these ideas using a model which arises in studies of cellular chemotaxis.
